Data-Driven Modelling \& \\ Forecasting of Chaotic Systems

### 2022-01-17 projectReport/chapter2.tex
Cleaner setup

\documentclass[a4paper,12pt,twoside]{report}
% \documentclass[a4paper,12pt,twoside]{book}
\pagestyle{headings}
\usepackage[utf8]{inputenc}
\usepackage{filecontents}
\usepackage[many]{tcolorbox}
\usepackage{graphicx}
\usepackage{harpoon}
%\usepackage{apacite} 
\usepackage{xcolor}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mdframed}
%\usepackage[authoryear]{natbib}
\usepackage{hyperref, url}
\usepackage[show]{ed}
\usepackage{subfigure}
\usepackage[ntheorem]{empheq} 
\usepackage{amsthm, amssymb, amsfonts, latexsym}
% \usepackage{pst-all}

\usepackage{enumitem}
\usepackage{multirow}
\usepackage{longtable}
\usepackage[export]{adjustbox}
% $f\arrowvert_A$ To restrict functions

\setlength{\parskip}{1.7em}
\setlength{\parindent}{0em}

%\setlength{\baselineskip}{20pt}

\setlength{\textwidth}{429.75499pt}
\setlength{\textheight}{643.20255pt}
\setlength{\oddsidemargin}{5 mm}
\setlength{\evensidemargin}{5 mm}
\setlength{\topmargin}{0 mm}
\setlength{\headsep}{0 mm}
\setlength{\headheight}{0 mm}

\usepackage{pst-node}

\newcommand{\citealtt}[1]{\citeauthor{#1},\citeyear{#1}}
\newcommand{\myycite}[1]{\citep{#1}}

\mathchardef\mhyphen="2D

%\newcommand{\cev}[1]{\reflectbox{\ensuremath{\vec{\reflectbox{\ensuremath{#1}}}}}}

\usepackage{pst-node,graphicx,pst-blur}
%\uspackage{auto-pst-pdf}
%\usepackage{tikz-cd}

\makeatletter
\DeclareRobustCommand{\cev}[1]{%
  \mathpalette\do@cev{#1}%
}
\newcommand{\do@cev}[2]{%
  \fix@cev{#1}{+}%
  \reflectbox{$\m@th#1\vec{\reflectbox{$\fix@cev{#1}{-}\m@th#1#2\fix@cev{#1}{+}$}}$}%
  \fix@cev{#1}{-}%
}
\newcommand{\fix@cev}[2]{%
  \ifx#1\displaystyle
    \mkern#20mu
  \else
    \ifx#1\textstyle
      \mkern#20mu
    \else
      \ifx#1\scriptstyle
        \mkern#26mu
      \else
        \mkern#26mu
      \fi
    \fi
  \fi
}

\makeatother

\newcommand{\xcm}{\epsfxsize=3.1cm}
\newcommand{\fig}[1]{\epsfbox}
% \newcommand{\bp}{\begin{minipage}{3.1cm}}
% \newcommand{\ep}{\end{minipage}}

\newtheorem{Definition}{Definition}[]
%\newmdtheoremenv{Theorem}{Theorem}[]
\newtheorem{Theorem}{Theorem}[]
\newtheorem{Lemma}{Lemma}[]
\newtheorem{Proposition}{Proposition}[]
\newtheorem{Corollary}{Corollary}[]
\newtheorem{Remark}{Remark}[]
\newtheorem{Example}{Example}[]

%Shortcut symbols
\newcommand{\Ftheta}{\ensuremath{F_\theta}}

\makeatletter 
\renewcommand{\thefigure}{\@arabic\c@figure}
\makeatother
\usepackage{xr}

%%%% PNAS GRAPHICS SETUP- FROM THE PNAS STYLE LATEX FILE
\RequirePackage{graphicx,xcolor}
\RequirePackage{colortbl}
\RequirePackage{booktabs}
\RequirePackage{algorithm}
\RequirePackage[noend]{algpseudocode}
\RequirePackage{changepage}
\RequirePackage[twoside,%
				letterpaper,includeheadfoot,%
				layoutsize={8.125in,10.875in},%
                layouthoffset=0.1875in,%
                layoutvoffset=0.0625in,%
                left=38.5pt,%
                right=43pt,%
                top=43pt,% 10pt provided by headsep
                bottom=32pt,%
                headheight=0pt,% No Header
                headsep=30pt,%
                footskip=25pt,
                marginparwidth=38pt]{geometry}
\RequirePackage[labelfont={bf,sf},%
                labelsep=period,%
                figurename=Fig. ]{caption}
\setlength{\columnsep}{13.5pt} % Distance between the two columns of text
\setlength{\parindent}{12pt} % Paragraph indent

%% Figure caption style
\DeclareCaptionFormat{pnasformat}{\normalfont\sffamily\fontsize{7}{9}\selectfont#1#2#3}
\captionsetup*{format=pnasformat}

%%% GREYBOX AROUND FIG
\definecolor{lightgray}{gray}{0.95}
\newcommand\greybox[1]{%
  \vskip\baselineskip%
  \par\noindent\colorbox{lightgray}{%
    \begin{minipage}{\textwidth}#1\end{minipage}%
  }%
  \vskip\baselineskip%
}

\newtcolorbox{paperbox}[1][]{
    enhanced,
    colback=white,
    boxrule=0pt,
    boxsep=0pt,
    arc=0mm,
    width=0.8\linewidth,
    fuzzy shadow={0mm}{-4pt}{-4pt}{1mm}{black!30!white},
    #1
}

% \externaldocument{Supp_Koopman-0808}
\begin{document}

\title{Data-Driven Modelling \& \\ Forecasting of Chaotic Systems}

\author{B.L. Nortier} 
\date{January, 2022}
\date{}
\maketitle

\begin{abstract}
  It is a well-established practice to collect measurements from underlying artificial, natural, and physical systems through observables in order to realise a dynamical model with greater descriptive power. 
  Many data-driven approaches to forecast the future evolution of systems exist. However, many of these methods rely on an initial understanding of the system and we often do not fully understand the many complexities of more intricate systems that we interact with on a daily basis.

  In this project, the base assumption is that data is originates from  some discrete-time dynamical system, particularly one of a chaotic nature but no assumptions are made pertaining to a specific dynamical system.
  A general learning problem is formulated and a well-known approach to the problem, alongside its practical limitations, is then considered.
  The notion of driven dynamical systems is introduced, and we subsequently establish the theoretical existence of a dynamical system possessing certain properties which guarantees it to be topologically conjugate (`dynamically equivalent`) to the dynamics of the underlying system from which measurements were originally taken. 
  Our methodology is implemented practically by making use of recurrent neural networks, a form of deep learning, to learn a map topologically equivalent to the underlying system that forecasts future states.
  In so doing we present the workability of an approach to obtain long-term topological and statistically consistent predictions for simple physical systems as well as a some chaotic attractors.
  \end{abstract}

  \tableofcontents

\chapter{Introduction}\label{ch1}

Experimenting with biological, physical, and artificial systems to generate a more informative dynamical model is a well-established practice in modern science. Traditional methods for modelling physical systems are based on laws of physics that are based either on empirical relationships or intuition. For systems that evolve with time, physical laws yield mathematical equations that govern how the quantities evolve with time. However, our world is much more complex than that which can be distilled into elegant equations. We do not fully understand many of the more complex systems, nor do they provide us with a good physical intuition of the underlying principles governing system dynamics. To complicate this, the underlying systems we observe often display sensitive dependence on initial conditions despite having highly similar initial conditions. This is because differing orbits diverge quickly and to such an extent that it becomes seemingly impossible to retrace their steps back to the original conditions. Even the presence of usually `negligible' computational noise or measurement error renders long-term pointwise prediction infeasible. We encounter many difficultiesin the process of modelling such systems:
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item One may not have access to the complete states of the systems
  \item The system may be described by functions that behave wildly because their graphs have a wild oscillatory behaviour, i.e., a large functional complexity.
\end{enumerate}

Models derived primarily from data can be classified into three categories: 
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item  Interpretable models (i.e., they establish relationships between internal physical mechanisms), 
  \item Partially interpretable models capturing some modes of the dynamics, 
  \item Non-interpretable models, defined as such mainly due to the fact that they are defined on a different phase space that is usually high-dimensional. 
\end{enumerate}

Examples in the literature attempting to forecast data from such systems have tried many different approaches with differing degrees of success. 
When we have complete access to states of the system, an ordinary differential equation can be obtained from data (\cite{brunton2016discovering, champion2019data} and \cite{small2002modeling,xu2006modeling}) wherein one could approximate the vector field by a library of functions to obtain interpretable models.

Recent partially interpretable models available in the literature have been based on the Koopman operator (see~\cite{koopman1932dynamical,budivsic2012applied}) to employ observables mapping the data onto a higher-dimensional space. This makes the dynamics in the higher-dimensional space more amenable for approximation by a linear transformation. Such methods do not guarantee exact reconstructions for nonlinear models, and in practice provide poor long-term consistency for a large class of chaotic dynamical systems \cite{Supp}.

The non-interpretable models include the delay embedding and the machine learning algorithms. For example, one could learn a system conjugate to the underlying system by applying the Takens delay embedding method of delay-coordinates~\cite{takens1981detecting} when one has "good" observations from the system; this learnt system could then be used to forecast the observed data. 
Takens delay embedding theorem~\cite{takens1981detecting} and its various generalisations (see,~\cite{sauer1991embedology, stark1999delay, gutman2018embedding}) establishes the learnability of a system constructed by concatenating sufficiently large previous time-series observations of a dynamical system into a vector (called delay coordinates). This then confirms the existence of a map on the space of delay coordinates equivalent (or topologically conjugate) to the underlying map from which the observed time-series was first obtained. Although topological conjugacy guarantees an alternate representation of the underlying system, the quality still depends on numerous parameters, making the comprehension of the dynamics unreliable (see, [?]). One reason for this fragility is that the embedded attractor in the reconstruction space is not always an attractor of the map learnt in the reconstruction space, despite unquestionably being an invariant set. When the embedded object is not explicitly known to be an attractor (as explained in Chapter~ \ref{ch3}), it can cause predictions to fail.

Practically, the application of Takens embedding involves learning a map through some technique, and consequently one wishes that these would have low functional complexity\cite{manjunath2021universal}, i.e., functions with fewer oscillatory graphs. On the other hand, pure machine learning methodology processes temporal information (like the echo state networks~\cite{Manju_ESP, Manju_IEEE, grigoryeva2018echo}) by mapping data onto a higher dimensional space for further processing. Although they perform  well on forecasting some dynamical systems, they fail completely on others (\ednote{Ask for resources here}) as there is often no guarantee that the right function was learnt during training.

This project involves implementing and analysing non-interpretable models that can guarantee exact reconstruction. 
The project work concerns the study and implementation of a method (\cite{manjunath2021universal}) that incorporates learning a function by mapping the data on to a higher dimensional space using what is called a driven dynamical system (See Chapter~ref{ch.4}). With a clear understanding of how the data is mapped onto the higher dimensional space, the method then permits the learning of a dynamical system topologically conjugate to that of the underlying system. Instead of linear regression as in the training of echo state networks, deep learning methods are employed to learn the correct function. With slight modifications to the implementation in the paper (\cite{manjunath2021universal}), we show that one can construct accurate non-interpretable models with the ability to reconstruct attractors from more hard-to-forecast systems like the double pendulum. (The forecasting of the time-series from a double pendulum has not been reported before.) Moreover, we also demonstrate that long-term statistical consistency is preserved.

\emph{By solidifying the mathematical underpinnings of our theory, we hope to guarantee the ability to construct models with predictive power ranging from molecular biology to neuroscience. in the near future. (We can modify this sentence when the report is completed.)}

The more intricate details of proofs are referred to where relevant throughout. This report is organised as follows: 
\newline In Chapter~\ref{ch2}, we recall the definition of a discrete-time dynamical system, how a discrete-system arises from a flow of an ODE and then proceed to define the inverse-limit space and topological conjugacy of autonomous systems. 
\newline Chapter~\ref{ch3} introduces the problem of forecasting dynamical systems, states the Takens delay embedding theorem, and discusses various issues faced while forecasting. 
\newline In Chapter~\ref{ch4}, a driven dynamical system is defined and discuss the properties of a specific class of driven dynamical systems that we make use of in this project.
\newline Finally, in Chapter~\ref{ch5} we show the implementation of these forecasting methods, and conclusions are provided in Chapter~\ref{ch6}.

\chapter{Discrete-time Dynamical Systems}\label{ch2}

@@

  \begin{figure}[ht]
    \includegraphics[scale=0.74]{_logistic_sdic.eps}
        \centering
        \captionof{figure}{Figure generated for Logistic Map $x_{n+1}=rx_n(1-x_n)$ with $r=4$ to exhibit the presence of SDIC. Plotted are the values $x_n$ against time $n$ for the the first 50 step with initial values $x_0=0.2$ (blue) and $x_0=0.20001$ (red). Initially the two trajectories overlap, but they diverge completely at $n=12$ whereafter they follow distinct paths.}
        \label{fig:log_sdic}
      \end{figure}
  \ednote{B: Caption and figure adjusted}
  % For an intuitive sense of the fact that the LM exhibits SDIC, consider again the definition and substitute specific values. Let $\delta=0.05$ and for purposes of illustration consider the value $u=0.4$. The green band in~\ref{fig:log_sdic} shades the region $[u-\delta, u+\delta]$ and we can see that regardless the value $v$  and in every neighborhood of $u \in U$ there exists a $v\in{U}$ and an integer $N:=N{(u,v)}\in\mathbb{Z}$ such that $d(T^Nu,T^Nv)>\delta$
  % specific point, say $u=0.4$. Draw a green neighbourhood $[0.35,0.45]$ of diameters $\delta=0.02$ around the point. It is easy to see that every neighbourhood around $u=0.4$ will contain some $v_1, v_2\in[0,1]$ and natural number $N_1, N_2 \in\mathbb{N}$ respectively s.t. the distance $d(T^Nu_1,T^Nv)>\delta=0.05$ . 
\end{Example}

@@

\begin{figure}[ht]
  \centering
\includegraphics[scale=0.8]{_dp_sdic.eps}
    \captionof{figure}{Three double pendulum heads with equal angular velocity and initial angles differing by 0,025 radians}\label{fig:dp_sdic}
  \end{figure}

@@

  \begin{figure}[ht]
    \includegraphics[scale=0.6]{_tentmap_2.eps}
        \centering
        \captionof{figure}{Graph of the $T^2$ map with bases at $x=0$, $x=\frac{1}{2}$ and $x=1$ respectively.}\label{fig:T2tentmap}
  \end{figure}

@@

\chapter{The Learning Problem}\label{ch3}

In this chapter, we acquaint ourselves with the question of forecasting dynamical systems with unknown underlying structures, state and discuss the shortcomings of the Takens delay embedding theorem and discuss various issues faced while forecasting.

Consider a relatively simple learning problem: 
Given the sequence $(u_0, u_1, \ldots, u_m)$ for $m\in\mathbb{Z}$, a finite segment of an orbit of the map $T$ where $T$ is defined by the update equation $u_{n+1} = Tu_n$, forecast the values $u_{m+1}, u_{m+1}$ where the map $T$ is unknown, given that $u_m$.

A practical example of this would be the subsequent scenario: given the time-sequential coordinates of an object moving in space, predict the future positions of that object. We are very rarely, if at all, presented with a problem where the entire state information is available to us in the form $u_n$ at some timestep $n\in\mathbb{N}$. Consider then a more complicated learning problem:

Suppose we only have the observations $\theta(u_0), \theta(u_1), \ldots, \theta(u_m)$ of the true system states $u$ in an unknown dynamical system $(U,T)$ and we wish to predict the values $\theta(u_{m+1}), \theta(u_{m+2})$ and so forth.

First we establish the method of Takens delay embedding. 
In this method, one considers a discrete-time dynamical system defined on a 'nice' space (a smooth manifold)\ednote{B: Should this be further explained?} that can be obtained as time-$K$ map of a continuous time-dynamical system, concepts formalised below.
We make observations from such a system, i.e., we observe the evolution of $\theta(w_0)$ (where $w_0$ represents the initial state) by examining  a finite set of values of $u_n :=  \theta(w_n) = \theta(Tw_{n-1})$. The sequence $\{u_n\}$ represents a scalar time-series and intuitively $\theta$ represents a probe inserted into a bigger system which is itself only measuring/extracting a small part of the greater system state at time $t=n$. 
Consider, for example, a thermometer erected to measure the ambient temperature in a local village. This measurement function, the thermometer, is capturing only a single aspect, the temperature, of a much grander dynamical system entailing the current weather of the surrounding area. Even more than that though- it is measuring a miniscule part of the global weather system.

%Given the sequence $\{u_1, u_2, \ldots, u_n}$, we define the variable \newline $y_n:=[\theta_n, \theta_{n-\tau}, \ldots, \theta_{n-(2d-2)\tau}, \theta_{n-(2d-1)\tau}]^{T}\in\mathbb{R}_{d}\times{1}$ where $\tau$ represents the lag and $d$ the dimension of the attractor. (We shall return to these terms shortly and their discussion may be put on hold for the time being).

However, the actual state $u$ of a system is seldom, if ever, fully known. the most one can do is to insert probes into a system to obtain partial information through the measurements taken. Moreover, the process of taking a measurement itself introduces two additional aspects which complicate the problem: \ednote{B: Perhaps this edit makes more sense? M: Maybe replace difficulty by saying that we need to consider two aspects. }
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item A series of measurements over a specific time interval is inherently a discretisation procedure of the underlying continuous-time dynamical system. (Hence why we restricted our attention to discrete-time systems in the preceding section)
  \item A probe will never be entirely accurate, and so the act of measurement introduces a certain measure of numerical noise/inaccuracy.
\end{enumerate}

From here we construct a multidimensional observable using the method of stacking previous observations, i.e., we create delay-coordinate map defined by
$\Phi_{k,\theta}(w) := (\theta(T^{-k}w)\ldots,\theta(T^{-1}w),\theta(w))$.  
The concept of an observable is understood in the sense of an observable as originally introduced by~\cite{takens1981detecting, genericObservableAeyels} to refer to a probe or measurement function inserted into the system. 
Takens' theorem, in essence, refers to the fact that for a sufficiently large $k$, we can define a dynamical system on the space $\mathbb{R}^{k+1}$ with states $\Phi_{k,\theta}(w)$, $\Phi_{k,\theta}(Tw)$, $\Phi_{k,\theta}(T^2w)$, etc. This dynamical system is topologically conjugate to the unknown underlying system $(W,T)$. We recall the Takens delay embedding theorem next.

%We may then ask ourselves the question: Is it in any way possible to retain information about the system state $x(t)$ in this temporal data-series $\theta(x(t))$? The answer is easily yes if T is known. However, neither this, nor even the exact function $\theta$ is available.

\section{Takens Embedding Theorem}\label{sect_Takens}

We define first the concepts of homeomorphism and embedding:
\begin{Definition}\rm
  [\bf {Homeomorphism}]\label{Dfn_homeo}\rm
  A homeomorphism is a function $f:Z\rightarrow Y$ between two topological spaces $Z$ and $Y$ that is continuous, bijective and has a continuous inverse. 
\end{Definition}

\begin{Definition}
  [\bf {Embedding}]\label{Dfn_embed}\rm
  Consider a homeomorphism $f:Z\rightarrow Y$ for $Y\subset X$. $Z$ is said to be embedded in $X$ by $f$.
\end{Definition}

Takens Theorem states a result establishing a relationship between the observed and underlying dynamical systems by showing that the concatenation of a sufficiently large number of previous observations into a vector will, under certain conditions, generate a map between the vectors from the respective systems.  We formulate the theorem from \cite{takens1981detecting}.

\begin{Theorem} 
	[\bf Takens Embedding Theorem (adopted from \cite{takens1981detecting}] \label{Thm_Takens}
         Let $W$ be a compact manifold of dimension $m$, and $d\ge m$ so that $2d$ is an integer. It is a 
            generic property for the pair $(T, \theta)$,  where $T:W \to W$ is
            a smooth diffeomorphism, and $\theta:W \to \mathbb{R}$ a smooth function, the map $\Phi_{2d,\theta}:W \to \mathbb{R}^{2d+1}$ defined on $W$ by 
            $\Phi_{2d,\theta}(w) := (\theta(T^{-2d}w)\ldots,\theta(T^{-1}w),\theta(w))$
            is a diffeomorphic embedding; by `smooth' we mean at least $C^2$. Consequently, there exists a map $F_\theta: \Phi_{2d,\theta}(W) \to \Phi_{2d,\theta}(W)$ defined by $$F_\theta: (\theta(T^{-2d}w),\ldots,\theta(T^{-1}w),\theta(w)) \mapsto 
            (\theta(T^{-2d+1}w),\ldots,\theta(w),\theta(Tw))$$
           so that $(W,T)$ is topologically conjugate to 
            $(\Phi_{2d,\theta}(W), F_\theta)$.    
\end{Theorem}

By generic we mean a residual set on a certain topology on appropriate spaces of functions (that we do not describe here) \ednote{B: Where can I find a citation for this?}. 
By $C^2$, we make reference to a twice-differentiable function  with a continuous second derivative. In our scenario (and this remains important throughout) the input space $U$ is considered the attractor.~\label{attractor_U}

Below we reproduce Figure 1 from \cite{Supp}.

%Some explanation is beneficial. Takens establishes a delay-coordinate map $\Ftheta$ defined in \ref{eqn_takens} for $T$ the flow (\ref{defn_flow}) defined by the update equation  $Tu_n=u_{n+1}$ and $theta$ our measurement function. When confining a dynamical system to the manifold U (in which the underlying system is contained), Takens showed that if certain smoothness conditions are satisfied on $T$ and $\theta$, then the delay-coordinate map $\Ftheta$ embeds $U$ in the reconstruction space $R^{2d+1}$ \textbf{we haven't used the term reconstruction space yet, so this is falling out of the blue}, for almost every choice of measurement function $\theta$, the observable. By almost every choice here, we mean that we exclude some observations which do not give any information. To see a more precise discussion, see \textbf{xxx} for we do not concern ourselves with a further elaboration on this report.

%Alternatively we may also make the statement: Takens' Embedding Theorem guarantees that almost all dynamical systems can be reconstructed from just one noiseless observation sequence \textbf{this isn't obvious from the theorem's formulation, is it?} i.e. for a great number of possible observation functions $\theta$, $\Ftheta$ preserves the topology of U. \textbf{'preserving the topology' is not something we've touched on before. Perhaps I should add to previous discussion so logic flows seamlessly here}.

\begin{figure}[ht]
  \includegraphics[scale=0.25]{_takensmap.eps}
  \centering
  \captionof{figure}{Schematic of Takens Embedding Theorem's usage and impact of iterates slipping from the attractor when the approximation $\tilde{F}_\theta$ of $F_\theta$ is learnt.}
  \label{fig:takensmap}
\end{figure}

If we can learn the map $F_\theta$, from sufficient number of data points of $\{\phi_{2d,\theta}(w_n)\}$, then we also know how to forecast how $\theta(w_n)$.

%Once again we pause and consider the graph to consolidate our understanding of the conjugacy (\textbf{are we being too simplistic?}). If we are in state $u$ and then evolve towards state $Tu$ before embedding into $\mathbb{R}^{2m-1}$, then this is equivalent to first embedding into $\mathbb{R}^{2m-1}$ and then evolving through the map $\Ftheta$, i.e. $\Phi_{2d,\theta}\circ{F}_{\theta} = T\circ\Phi_{2d,\theta}$. The map $F_theta$ is a homeomorphism as well.

%Thus information about U can be retained in the time series' (or observation's) output. By preserving the topology on the manifold $U$ in the reconstruction space $X$ (\textbf{not discussed before}), we are also guaranteed the preservation of topological invariants of the manifold, of which dimensionality is one such invariant. (We note here that dimensionality will again be considered later in this project).

\section{Practical Limitations to Takens' Theorem}
Takens' Embedding Theorem is a powerful result, and it provides compelling reason to believe that one could conceivably reconstruct a system conjugate accurately to the underlying system. 
Nonetheless, it does present some severe practical limitations:
\vspace{-5mm}
\begin{enumerate}
\item Even supposing that we could indeed find $F_\theta$ , our \emph{approximation} of  $F_\theta$ is a map from a larger set $\mathbb{R}^{2d+1}$ containing the embedded attractor. There are, however, no theoretical guarantees that $F_\theta$ will retain  $V=\Phi_{2d,\theta}(W)$  as an attractor although $W$ could itself be an attractor.
\item Takens theorem is stated only for noiseless observations. Due to noise $\epsilon_n$ the delay vector $\Phi_{2d,\theta}(w_n) + \epsilon_n$  may lie outside $V$.Furthermore, due to the chaotic nature of the underlying system (i.e. the fact that it has SIDC), the evolution of $\Phi_{2d,\theta}(w_n) + \epsilon_n$ under the map $F_\theta$ could move out of $V$ completely. This problem can be overcome  by using a driven dynamical system with some properties, and we discuss this in Chapter\ref{ch3}. 
\end{enumerate}

% The Takens embedding theorem does not guarantee global dissipativity.
% \ednote{B:Links with~\ref{subs_LearnGamma} in discussion of dissipativity of $\Gamma$.  Sir, I'm not sure how to write this here - I think it a beneficial point to make so as to set the stage for the later discussion in ~\ref{subs_LearnGamma}, but I am unsure there as well.}

Though not a uniquely related to the Takens embedding theorem, we do opt to make mention of another requirement for our work. 
The methods described here are applicable for data originating from a surjective map. 
This follows from the fact that when $T$ maps the space $U$ into a proper subset $B$ of$ $U, there exists some point in $U - A$ that does not have a preimage and we conclude that $T$ is not surjective. 
An example of this would be the case when a system displays energy-loss. Consider, for instance, consider the dying oscillations for a damped pendulum. (An account of greater depth  is provided in~\ref{ch5})

This requirement is not overly restrictive, as a great number of chaotic systems do have a surjective map. (Cite) \ednote{B: I remember you saying this in one of our sessions, but don't have a source for it. Do you have a specific reference?}

\chapter{Driven Dynamical Systems to Forecast Problems}\label{ch4}

In this chapter, we discuss results on the mapping of temporal data obtained from a discrete-dynamical system onto a different space through the notion of a driven dynamical system. 
We also consider the conditions a driven system should possess to avoid adding distortion to its state-space representation. We then describe the \emph{single-delay dynamics} (SDD) of the system and finally recall conditions on the driven system so that the SDD are conjugate (or at least semi-conjugate) to the underlying system. 
The SDD can then be used to forecast and reconstruct the underlying system.

\section{Nonautonomous and Driven Dynamical Systems}

\begin{Definition}
  [\bf Nonautonomous Dynamical System (adopted from~\cite{Manju_ESP}))] \label{Dfn_NDS}\rm
  A nonautonomous dynamical system (NDS) on a space $X$ is simply a dynamical system comprising a family of maps $\{g_n\}_{n \in \mathbb{Z}}$ where $g_n(\ldotp):=g(u_n,\ldotp):X\to{X}$ is continuous. Each $g_n$ arises as a consequence of an input $u_n$ from the input space $U$, a topological space. 
\end{Definition}

We immediately recall the figure first defined in \eqref{eqn_conjugacy}  and remind ourselves of our ultimate goal to learn a system that is conjugate to an underlying or unknown dynamical system $(U,T)$ by making use of the measurements obtained from  the unknown system.

%#####################################
\begin{center}
\psset{arrows=->, arrowinset=0.25, linewidth=0.6pt, nodesep=3pt, labelsep=2pt, rowsep=0.7cm, colsep = 1.1cm, shortput =tablr}
\everypsbox{\scriptstyle}
\begin{psmatrix}
U & U\\%
V & V.
%%%
%  \ncline{1,1}{1,2}^{T} 
%  \ncline{1,1}{2,1} <{\phi}
%   \ncline{2,1}{2,2}^{S}
%  \ncline{1,2}{2,2} > {\phi}
\end{psmatrix}
%#####################################
\end{center}

To find a function $\phi$, the need now arises to consider a so-called driven dynamical system, a special case of a nonautonomous dynamical system. Two spaces are considered: input is taken from an input space $U$ and the state of a space $X$ is updated. (Previously, only the state was considered.)
We do this to mimic the true conditions. By accounting for an extrogenous input sequence in $U$ which can also influence the  system-state at a specific time-step, it produces more general models than an autonomous system.

\begin{Definition}
  [\bf Driven, Compactly Driven Dynamical System] \label{Dfn_DDS} \rm
A driven dynamical system comprises two topological metric spaces $U$, $X$ and a continuous function  $g:U\times{X}\to{X}$ where $g(u_n, x_n)=x_{n+1}$.
If the input space $U$ is compact, we refer to the system as compactly driven. 
\end{Definition}

The update equation $x_{n+1} = g(u_n,x_n)$ for $n \in\mathbb{Z}$, input $u_n$ from $U$ and state $x_n$ belonging to $X$ (a compact space) generate the dynamics on X. 
Abbreviated, we refer only to the \textit{driven system $g$}, where all other entities are understood implicitly.

Notably, a nonautonomous dynamical system can be generated from $U$; any input $\overline{u}$, a bi-infinite sequence from $U$, gives rise to the sequence of self-maps $\{g(u_n, \cdot)\}_{n\in\mathbb{Z}}$ contained in $X$.
Physically, one may think of this bi-infinite sequence as referring to a system that has been running for an extended period at the time of the first measurement taken from the system (alternatively, the first time a probe is inserted into the system to take an observation)

\begin{Definition}
  [\bf Entire Solution] \label{Dfn_Soln} \rm
  A sequence $\{x_n\}_{n\in\mathbb{Z}}\subset X$ is called an entire solution (or simply a solution) of the driven system  $g$ with input $\overline{u}$ when it satisfies 
  \[g(u_{n-1}, x_{n-1})=x_n\] for all $n\in\mathbb{Z}$
\end{Definition}

It is important to emphasise that a sequence $\{x_n\}$ satisfying the update equation above can only be a solution if $x_n\in{U}$ holds for all $n\in\mathbb{Z}$ and just for $n>0$. Consider the example below.

\begin{Example} \rm \label{ex_halfux}
  The only solution  $\{x_n\}_{n\in\mathbb{Z}}$ to the driven system  $g(u,x)=\frac{ux}{2}$, where $X=[0,1]$, $U=[0,1]$,  is the zero solution $x_n\equiv0$.
  To see this, consider any $x_n=a\in[0,1]$ where $a\neq{0}$.  Let $\overline{u}\in{U}$ be an non-zero constant sequence , say $u_n=0.5$. 
  The driven system may be rewritten as $x_{n-1}=\frac{2x_n}{u_{n-1}}=4x_n$ and the  iterates of $x_n$ in backward time will increase by a factor of 4 at each timestep. 
  Thus for some $m\leq{n}$,  $1<x_m$ i.e. $x_m\notin{X}$. So ${\{x_n\}}_{n\in\mathbb{Z}}$ is not a solution and it follows that the only possible solution is the zero solution.
\end{Example}
% \ednote{Nicely written! Thank you}

A system may also have multiple solutions, as evidenced in the example below.

\begin{Example}\label{Ex_exp} \rm 
  Consider the driven system $g(u,x)=x^2$ for $X=[0,1]$, $U=\mathbb{R}$. The system has an uncountable number of solutions, as there exists a solution for every $x\in{X}$ which also passes through the point $x$ and $lim_{n\to\infty}x_n=0$, $lim_{n\to{-}\infty}x_n=1$.  
 The proof is deferred to immediately after the next paragraph (see~\ref{rem_proofEx}).
\end{Example}

As the solutions to a driven system are considered an important entity, we next identify a subspace $X_U$ of $X$ that contains all possible solutions. To realize such a subspace of a driven system $g$, the concept of a reachable set is defined.

\begin{Definition}
  [\bf Reachable Set]\label{Dfn_ReachableSet}\rm
The reachable set of a driven system $g$ is exactly the union of all the elements of all the (entire) solutions, i.e., 
\[X_U :=\Big \{x \in X:  x = x_k \mbox{ where  $\{x_n\}$  is a solution for some  $\bar{u}$} \Big \}.\]
The set of all reachable states at a specific time $n$ for input $\overline{u}$ is denoted by $X_n(\overline{u})$
\end{Definition}

The reachable set itself can defined independently from $g$ possessing the ESP/USP. Note that $x\in{X_n(\overline{u})}$ if and only $g$ has a solution $\{x_k\}$ for $x_n=x$ and input $\overline{u}$, a result proved in nonautonomous dynamical systems literature~\cite{manjunath2014dynamics, manjunath2013echo}. \ednote{B: Adjusted sentence and cited}

The set $X_n(\overline{u})$ is precisely the set of points $x$ through which some entire solution $\Psi$ passes at time $n$. We can give an alternate formulation of the set $X_n(\overline{u})$. 
Now suppose, we for some fixed input 
$\overline{u}$, we define $g_i = g(u_i,\cdot)$ for all $i\in \mathbb{Z}$, then the set of states

\begin{equation} \label{eqn_association}
X_{n,i}(\overline{u}) := g_{n-1} \circ \cdots g_{i+1} \circ g_i(X).
\end{equation}

The set $X_{n,i}(\overline{u})$ being the image of a finite composition of
continuous maps is compact whenever $X$ is compact. Also $X_{n,i}(\overline{u})$ is
nonempty.  Further, $X_{n,i}(\overline{u}) \supset X_{n,i-1}$. Hence $\bigcap_{i<n}
X_{n,i}(\overline{u})$ is a nested intersection of closed nonempty subsets, and
whenever $X$ is compact, the intersection is nonempty. We say $g$ is a topological contraction if $\bigcap_{i<n}X_{n,i}(\overline{u})$ is a singleton subset of $X$ for each $\overline{u}$.

It turns out that $\bigcap_{i<n}X_{n,i}(\overline{u})$ is identically equal to $X_n((\overline{u})$ that we had defined earlier. This will lead to a result (for. see, \cite{manjunath2013echo}) that  \textit{$g$ being a topological contraction is equivalent to the existence of a exactly one entire solution.}

>>>>

@@

  \begin{figure}[ht]
    \includegraphics[scale=0.74]{Graphs/_logistic_sdic.eps}
        \centering
        \captionof{figure}{Figure generated for Logistic Map $x_{n+1}=rx_n(1-x_n)$ with $r=4$ to exhibit the presence of SDIC. Plotted are the values $x_n$ against time $n$ for the the first 50 step with initial values $x_0=0.2$ (blue) and $x_0=0.20001$ (red). Initially the two trajectories overlap, but they diverge completely at $n=12$ whereafter they follow distinct paths.}
        \label{fig:log_sdic}
      \end{figure}
  \ednote{B: Caption and figure adjusted}
  % For an intuitive sense of the fact that the LM exhibits SDIC, consider again the definition and substitute specific values. Let $\delta=0.05$ and for purposes of illustration consider the value $u=0.4$. The green band in~\ref{fig:log_sdic} shades the region $[u-\delta, u+\delta]$ and we can see that regardless the value $v$  and in every neighborhood of $u \in U$ there exists a $v\in{U}$ and an integer $N:=N{(u,v)}\in\mathbb{Z}$ such that $d(T^Nu,T^Nv)>\delta$
  % specific point, say $u=0.4$. Draw a green neighbourhood $[0.35,0.45]$ of diameters $\delta=0.02$ around the point. It is easy to see that every neighbourhood around $u=0.4$ will contain some $v_1, v_2\in[0,1]$ and natural number $N_1, N_2 \in\mathbb{N}$ respectively s.t. the distance $d(T^Nu_1,T^Nv)>\delta=0.05$ . 
\end{Example}

@@

\begin{figure}[ht]
  \centering
\includegraphics[scale=0.8]{Graphs/_dp_sdic.eps}
    \captionof{figure}{Three double pendulum heads with equal angular velocity and initial angles differing by 0,025 radians}\label{fig:dp_sdic}
  \end{figure}

@@

  \begin{figure}[ht]
    \includegraphics[scale=0.6]{Graphs/_tentmap_2.eps}
        \centering
        \captionof{figure}{Graph of the $T^2$ map with bases at $x=0$, $x=\frac{1}{2}$ and $x=1$ respectively.}\label{fig:T2tentmap}
  \end{figure}

@@
